---
title: "Risk assesment with quantrra"
author: "Pablo Gomez"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Risk assesment with quantrra}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```



# Basics

```{r loading packages}
# Libraries we will use
library(sf); library(dplyr)
library(quantrra)
```

## Sampling probability distributions

By default, R has several functions to sample a distribution and plot the results. For example, if we would like to sample a normal distribution with a mean of 5 and standard deviation of 0.12, we can use the following code:

```{r distribution sampling}
# sample the distribution and save it to an object
x <- rnorm(n = 100, # Number of observations to sample
           mean = 5, # Mean 
           sd = 0.12) # Standard deviation

# Plot the observations
hist(x)
```

In this lab we will be using the package `QuantRRA`, which is specifically developed for rapid risk assessment in R. The function `ra_sample()` support multiple distributions such as: Normal, binomial, uniform, pert, among others. The function requires two arguments:  
  
  - `x` which is a string (text) of the distribution and its parameters.  
  - `n` the number of observations we will sample.  

```{r QuantRRA distribution sampling}
n <- 100 # number of observations
d <- 'Normal(5, 0.12)' # Distribution to sample

x <- ra_sample(x = d, n = n) # Function to sample the distribution

# We can use the function plotDist() from the package QuantRRA to get a more familiar output:
ra_plotDist(x, # the values sampled
         main = 'Distribution of x') # A title for our plot
```



